---
title: 'Practical Machine Learning Project Report'
author: 'Ioannis Nennes'
output: html_document
---

## Background ##

Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement – a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website [here](http://groupware.les.inf.puc-rio.br/har). (see the section on the Weight Lifting Exercise Dataset). 

## Loading Data ## 

The training data for this project are available [here](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv). The test data are available [here](https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv). The data for this project come from this [source](http://groupware.les.inf.puc-rio.br/har). If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment. 

The above data has been downloaded to the 'data' directory. The following code loads the data and the necessary libraries.

```{r loading}
suppressMessages(library(knitr, warn.conflicts = FALSE, quietly=TRUE))

opts_chunk$set(
  cache=TRUE
  , dev='png'
  , dev.args=list(type="cairo")
  , fig.height=4
  ,dpi=300
  , warning=FALSE
  , message=FALSE
  , results="hide"
  , echo=TRUE
  , eval=TRUE
  , xtable.type="html")
knit_hooks$set(inline = function(x) {
  prettyNum(x, big.mark=",")
})

# Library the required dependencies
suppressMessages(library(caret, warn.conflicts = FALSE, quietly=TRUE))
suppressMessages(library(dplyr, warn.conflicts = FALSE, quietly=TRUE))
suppressMessages(library(randomForest, warn.conflicts = FALSE, quietly=TRUE))

#load raw data
setwd("C:/Users/John/Dropbox/Code/GitHub/ML-coursera")
training <- read.csv("data/pml-training.csv", header = TRUE, na.strings=c("NA","","#DIV/0!"))
testing  <- read.csv('data/pml-testing.csv', header = TRUE, na.strings=c("NA","","#DIV/0!"))
  
#set the classe feature type
training$classe <- as.factor(training$classe)
```

## Cleaning Data ##

After examining the data, we have identified a number of actions required to clean the dataset:

* There are a few columns where the majority of the values are NA. We will remove columns where there are more than 50% NA values.
* We wil also remove the first 7 columns, as they do not contain measurements.
* Finally, near zero variance features will be removed.

```{r cleaning}
# We will remove columns where there are more than 50% NA values.
na.rich.cols<- apply(training,2,function(x) {sum(is.na(x))});
training <- training[,which(na.rich.cols <  nrow(training)*0.5)];  

# We wil also remove the first 7 columns, as they do not contain measurements.
training<-training[,8:ncol(training)]

# Finally, near zero variance features will be removed.
nz.cols <- nearZeroVar(training, saveMetrics = TRUE)
training <- training[, nz.cols$nzv==FALSE]

```

## Creating the Cross Validation set ##

We will use 75% of the training data set for training our model and the remaining 25% for cross validating it.

```{r cross_validation}

set.seed(221)

training.idx = createDataPartition(training$classe, p = 3/4, list=FALSE)
training.part <- training[training.idx,];
testing.part <- training[-training.idx,];

```

## Model Comparison ##

We will try three models: Random forest ("rf"), boosted trees ("gbm") and linear discriminant analysis ("lda"). Based on the accuracy of the predictions, we will choose the optimal.

```{r models}
set.seed(222)

# We will try three models: Random forest ("rf"), boosted trees ("gbm") and linear discriminant analysis ("lda").
train.rf <- train(classe ~ ., method = 'rf' , data = training.part, 
             # add tuning par. mtry to avoid long lasting computations for the
             # ideal mtry; take the randomforest dft: sqrt(number of features)
             tuneGrid = data.frame(mtry = floor(sqrt(ncol(training.part)-1)))
            )   
train.gbm <-train(classe ~ ., method = 'gbm', data = training.part)
train.lda <-train(classe ~ ., method = 'lda', data = training.part) 

predict.rf <- predict(train.rf  , testing.part)
predict.gbm<- predict(train.gbm , testing.part)
predict.lda<- predict(train.lda , testing.part)
```

Random forest has an accuracy of `r confusionMatrix(predict.rf, testingPartition$classe)`, boosted trees have an accuracy of `r confusionMatrix(predict.gbm, testingPartition$classe)` and linear discriminant analysis has an accuracy of `r confusionMatrix(predict.lda, testingPartition$classe)`. Therefore we will use the Random Forest model.

## Tuning with Cross Validation ##

To avoid over-fitting, we will use cross-validation on the test set created inside the original training set.

```{r cv_tuning}
set.seed(223)

fit.control <- trainControl(method = "repeatedcv", number = 10, repeats = 10)
model.rf.cv <- train(classe ~ ., method="rf",  data=training.part, trControl = fit.control, 
                 # add tuning par. mtry to avoid long lasting computations for the
                 # ideal mtry; take the randomforest dft: sqrt(number of features)
                 tuneGrid = data.frame(mtry = floor(sqrt(ncol(training.part)-1)))
                )   

predict.rf.cv <<- predict(model.rf.cv , testing.part)
```

Tuning proved usefull, as the accuracy was improved from `r ` to `r `.

## Predicting on the test set ##

```{r test_predict}
predict.final <<- predict(model.rf.cv , testing)
```

We will store the predictions on the 'predictions' directory, using the code below:

```{r save_predict}
answers <- as.vector(predict.final)

pml_write_files = function(x) {
    n = length(x)
    for (i in 1:n) {
        filename = paste0("problem_id_", i, ".txt")
        write.table(x[i], file = filename, quote = FALSE, row.names = FALSE, 
            col.names = FALSE)
    }
}

pml_write_files(answers)
```





